---
title: "Evaluations of SVGs methods: mOB dataset"
author: "Lukas Weber"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
```


# Introduction

Evaluations of SVGs methods:

- mOB dataset
- using implementation of BRISC method in nnSVG package

```{r, message = FALSE}
library(SpatialExperiment)
library(STexampleData)
library(nnSVG)
library(dplyr)
library(tidyr)
library(ggplot2)
```


---

# Load data and preprocessing


## Load data

Load dataset from MERINGUE package and reshape into `SpatialExperiment` format.

```{r, message=FALSE}
library(MERINGUE)
```

```{r}
# load dataset from MERINGUE package
data(mOB)

names(mOB)
dim(mOB$counts)
mOB$counts[1000:1006, 1:6]
head(mOB$pos)
head(mOB$results)
str(mOB$annot)

# create SpatialExperiment
row_data <- data.frame(
  gene_name = rownames(mOB$counts)
)
res <- cbind(mOB$results, gene_name = rownames(mOB$results))
row_data <- DataFrame(left_join(row_data, res, by = "gene_name"))
head(row_data)

col_data <- data.frame(
  barcode_id = colnames(mOB$counts)
)
annot <- data.frame(
  barcode_id = names(mOB$annot), 
  layer_id = mOB$annot
)
col_data <- DataFrame(left_join(col_data, annot, by = "barcode_id"))
head(col_data)

spatial_coords <- as.matrix(mOB$pos[col_data$barcode_id, ])
head(spatial_coords)

stopifnot(nrow(spatial_coords) == nrow(col_data))
stopifnot(all(rownames(spatial_coords) == col_data$barcode_id))
stopifnot(all(colnames(mOB$counts) == col_data$barcode_id))

spe <- SpatialExperiment(
  assays = list(counts = mOB$counts), 
  rowData = row_data, 
  colData = col_data, 
  spatialCoords = spatial_coords
)

spe
```


## Preprocessing

Preprocessing steps: filter low-expressed genes, filter mitochondrial genes, normalization, calculate logcounts.

Using convenience function in `nnSVG` package.

```{r}
set.seed(123)
spe <- preprocessSVG(spe, in_tissue = FALSE)

dim(spe)
assayNames(spe)

logcounts(spe)[1:6, 1:6]
```


## Clustering

Clustering to identify major cell types, which can be provided as a covariate in the `nnSVG` model.

```{r}
set.seed(123)
spe <- clusterSVG(spe)

colData(spe)

# compare against ground truth labels
table(truth = colData(spe)$layer_id, 
      clusters = colData(spe)$label)
```


---

# Identify SVGs

## nnSVG: all spots

Using implementation of BRISC method in nnSVG package

Runtime: ~45 min with 4 cores on laptop

```{r}
# # run nnSVG across all spots
# runtime_nnSVG_all <- system.time({
#   spe_all <- nnSVG(spe, n_threads = 4)
# })
# 
# runtime_nnSVG_all
```

```{r}
# # save output
# file <- "../../outputs/spe_mOB_nnSVG_all.rds"
# saveRDS(spe_all, file = file)
```

```{r}
# alternatively: load saved results
spe_all <- readRDS("../../outputs/spe_mOB_nnSVG_all.rds")
```


## nnSVG: within cell types

Including covariates for cluster IDs representing cell types

Runtime: ~45 min with 4 cores on laptop

```{r}
# # build model matrix using cell type labels from clustering
# X <- model.matrix(~ colData(spe)$label)
# dim(X)
# head(X)
# stopifnot(nrow(X) == ncol(spe))
# 
# # run nnSVG taking cell types into account
# runtime_nnSVG_clusters <- system.time({
#   spe_clusters <- nnSVG(spe, x = X, n_threads = 4)
# })
```

```{r}
# # save output
# file <- "../../outputs/spe_mOB_nnSVG_clusters.rds"
# saveRDS(spe_clusters, file = file)
```

```{r}
# alternatively: load saved results
spe_clusters <- readRDS("../../outputs/spe_mOB_nnSVG_clusters.rds")
```


---

# Comparisons

## Top SVGs

```{r}
# all spots
rowData(spe_all)[rowData(spe_all)$rank <= 10, ]

# within cell types
rowData(spe_clusters)[rowData(spe_clusters)$rank <= 10, ]
```


## Plots

```{r, fig.width=4, fig.height=3}
# plot clusters
df <- as.data.frame(cbind(spatialCoords(spe), colData(spe)))

ggplot(df, aes(x = x, y = y, color = label)) + 
  geom_point(size = 2) + 
  coord_fixed() + 
  scale_y_reverse() + 
  ggtitle("Cluster labels") + 
  theme_bw() + 
  theme(panel.grid = element_blank(), 
        axis.title = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank())
```


```{r, fig.width=4, fig.height=3}
# top SVG across all spots
ix <- which(rowData(spe_all)$rank == 1)
ix_name <- rowData(spe_all)$gene_name[ix]

df <- as.data.frame(
  cbind(spatialCoords(spe_all), 
        expr = counts(spe_all)[ix, ]))

ggplot(df, aes(x = x, y = y, color = expr)) + 
  geom_point(size = 2) + 
  coord_fixed() + 
  scale_y_reverse() + 
  scale_color_gradient(low = "gray90", high = "red") + 
  ggtitle(ix_name) + 
  theme_bw() + 
  theme(panel.grid = element_blank(), 
        axis.title = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank())
```


```{r, fig.width=4, fig.height=3.5}
# top SVG within cell types
ix <- which(rowData(spe_clusters)$rank == 1)
ix_name <- rowData(spe_clusters)$gene_name[ix]

df <- as.data.frame(
  cbind(spatialCoords(spe_clusters), 
        expr = counts(spe_clusters)[ix, ]))

ggplot(df, aes(x = x, y = y, color = expr)) + 
  geom_point(size = 2) + 
  coord_fixed() + 
  scale_y_reverse() + 
  scale_color_gradient(low = "gray90", high = "red") + 
  ggtitle(ix_name) + 
  theme_bw() + 
  theme(panel.grid = element_blank(), 
        axis.title = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank())
```


---

# Session info

```{r}
sessionInfo()
```

