---
title: "Mean-variance relationships"
author: "Lukas Weber"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
```


# Introduction

Mean-variance relationship in Visium datasets

```{r, message = FALSE}
library(SpatialExperiment)
library(STexampleData)
library(matrixStats)
library(MASS)
library(dplyr)
library(tidyr)
library(ggplot2)
```


# Visium human DLPFC: layers 1 to 6

## Load data

Visium human DLPFC dataset, sample 151673

```{r}
# load dataset from STexampleData package
spe <- Visium_humanDLPFC()
dim(spe)
```


## Mean-variance plot

Select spots:

```{r}
# select spots over tissue
spe <- spe[, spatialData(spe)$in_tissue == 1]
dim(spe)

# select spots in layers 1 to 6
spe <- spe[, colData(spe)$ground_truth %in% paste0("Layer", 1:6)]
dim(spe)
```


Select genes:

```{r}
# remove genes with zero expression
ix_zeros <- rowSums(counts(spe)) == 0
spe <- spe[!ix_zeros, ]
dim(spe)
```


Calculate gene-level values (mean and variance):

```{r}
# calculate gene-level values
df <- data.frame(
  gene_id = rowData(spe)$gene_id, 
  gene_name = rowData(spe)$gene_name, 
  mean = rowMeans(as.matrix(counts(spe))), 
  observed = rowVars(as.matrix(counts(spe)))
)

dim(df)
head(df)
```


Calculate theoretical values for variance from Poisson distribution. Poisson distribution has `variance = mean`, so these are simply equal to the `mean`.

```{r}
# theoretical Poisson values
# (variance = mean)
df$poisson <- df$mean
```


Calculate theoretical values for variance from negative binomial distribution. This requires fitting values to estimate the dispersion parameter.

```{r}
# fit negative binomial to data (using MASS package)
gene_sums <- rowSums(counts(spe))
summary(glm.nb_out <- glm.nb(gene_sums ~ 1))

# estimated dispersion parameter
nb_disp <- glm.nb_out$theta
nb_disp

# theoretical negative binomial values
# (variance = mean + mean^2 * disp)
df$neg_binom <- df$mean + (df$mean^2) * nb_disp
```


Create plot:

```{r, fig.width=5.5, fig.height=4}
# reshape data frame
df <- pivot_longer(
  df, 
  cols = c("observed", "poisson", "neg_binom"), 
  names_to = "method", 
  values_to = "variance"
)

df$method <- factor(df$method, levels = c("observed", "poisson", "neg_binom"))
head(df)

# create plot
ggplot(as.data.frame(df), aes(x = mean, y = variance, 
                              group = method, color = method)) + 
  geom_point(pch = 1, size = 0.8) + 
  scale_x_log10() + 
  scale_y_log10() + 
  scale_color_manual(values = c("black", "blue", "red")) + 
  ggtitle("Visium human DLPFC: layers 1 to 6") + 
  theme_bw()
```


---

# Visium human DLPFC: layer 6

## Load data

Visium human DLPFC dataset, sample 151673

```{r}
# load dataset from STexampleData package
spe <- Visium_humanDLPFC()
dim(spe)
```


## Mean-variance plot

Select spots:

```{r}
# select spots over tissue
spe <- spe[, spatialData(spe)$in_tissue == 1]
dim(spe)

# select spots in layer 6
spe <- spe[, colData(spe)$ground_truth %in% "Layer6"]
dim(spe)
```


Select genes:

```{r}
# remove genes with zero expression
ix_zeros <- rowSums(counts(spe)) == 0
spe <- spe[!ix_zeros, ]
dim(spe)
```


Calculate gene-level values (mean and variance):

```{r}
# calculate gene-level values
df <- data.frame(
  gene_id = rowData(spe)$gene_id, 
  gene_name = rowData(spe)$gene_name, 
  mean = rowMeans(as.matrix(counts(spe))), 
  observed = rowVars(as.matrix(counts(spe)))
)

dim(df)
head(df)
```


Calculate theoretical values for variance from Poisson distribution. Poisson distribution has `variance = mean`, so these are simply equal to the `mean`.

```{r}
# theoretical Poisson values
# (variance = mean)
df$poisson <- df$mean
```


Calculate theoretical values for variance from negative binomial distribution. This requires fitting values to estimate the dispersion parameter.

```{r}
# fit negative binomial to data (using MASS package)
gene_sums <- rowSums(counts(spe))
summary(glm.nb_out <- glm.nb(gene_sums ~ 1))

# estimated dispersion parameter
nb_disp <- glm.nb_out$theta
nb_disp

# theoretical negative binomial values
# (variance = mean + mean^2 * disp)
df$neg_binom <- df$mean + (df$mean^2) * nb_disp
```


Create plot:

```{r, fig.width=5.5, fig.height=4}
# reshape data frame
df <- pivot_longer(
  df, 
  cols = c("observed", "poisson", "neg_binom"), 
  names_to = "method", 
  values_to = "variance"
)

df$method <- factor(df$method, levels = c("observed", "poisson", "neg_binom"))
head(df)

# create plot
ggplot(as.data.frame(df), aes(x = mean, y = variance, 
                              group = method, color = method)) + 
  geom_point(pch = 1, size = 0.8) + 
  scale_x_log10() + 
  scale_y_log10() + 
  scale_color_manual(values = c("black", "blue", "red")) + 
  ggtitle("Visium human DLPFC: layer 6") + 
  theme_bw()
```


---

# Visium mouse coronal

## Load data

Visium mouse coronal dataset

```{r}
# load dataset from STexampleData package
spe <- Visium_mouseCoronal()
dim(spe)
```


## Mean-variance plot

Select spots:

```{r}
# select spots over tissue
spe <- spe[, spatialData(spe)$in_tissue == 1]
dim(spe)
```


Select genes:

```{r}
# remove genes with zero expression
ix_zeros <- rowSums(counts(spe)) == 0
spe <- spe[!ix_zeros, ]
dim(spe)
```


Calculate gene-level values (mean and variance):

```{r}
# calculate gene-level values
df <- data.frame(
  gene_id = rowData(spe)$gene_id, 
  gene_name = rowData(spe)$gene_name, 
  mean = rowMeans(as.matrix(counts(spe))), 
  observed = rowVars(as.matrix(counts(spe)))
)

dim(df)
head(df)
```


Calculate theoretical values for variance from Poisson distribution. Poisson distribution has `variance = mean`, so these are simply equal to the `mean`.

```{r}
# theoretical Poisson values
# (variance = mean)
df$poisson <- df$mean
```


Calculate theoretical values for variance from negative binomial distribution. This requires fitting values to estimate the dispersion parameter.

```{r}
# fit negative binomial to data (using MASS package)
gene_sums <- rowSums(counts(spe))
summary(glm.nb_out <- glm.nb(gene_sums ~ 1))

# estimated dispersion parameter
nb_disp <- glm.nb_out$theta
nb_disp

# theoretical negative binomial values
# (variance = mean + mean^2 * disp)
df$neg_binom <- df$mean + (df$mean^2) * nb_disp
```


Create plot:

```{r, fig.width=5.5, fig.height=4}
# reshape data frame
df <- pivot_longer(
  df, 
  cols = c("observed", "poisson", "neg_binom"), 
  names_to = "method", 
  values_to = "variance"
)

df$method <- factor(df$method, levels = c("observed", "poisson", "neg_binom"))
head(df)

# create plot
ggplot(as.data.frame(df), aes(x = mean, y = variance, 
                              group = method, color = method)) + 
  geom_point(pch = 1, size = 0.8) + 
  scale_x_log10() + 
  scale_y_log10() + 
  scale_color_manual(values = c("black", "blue", "red")) + 
  ggtitle("Visium mouse coronal") + 
  theme_bw()
```


---

# Session info

```{r}
sessionInfo()
```

