---
title: "Comparison of SVGs methods: BRISC extensions"
author: "Lukas Weber"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Introduction

Analyses for SVGs: comparison of methods including BRISC, GpGp, HVGs, using spatzli wrappers, human DLPFC dataset.

```{r, eval = FALSE}
# module load conda_R/4.1.x
# R
```

```{r, message = FALSE}
library(SpatialExperiment)
library(STexampleData)
library(spatzli)
library(scran)
library(SPARK)
library(dplyr)
library(ggplot2)
library(tidyr)
```


# Load data

```{r}
spe <- Visium_humanDLPFC()
dim(spe)
assayNames(spe)
```


# Preprocessing

Preprocessing steps: filter low-expressed genes, filter mitochondrial genes, normalization, and logcounts. Using convenience function in `spatzli` package.

```{r}
spe <- preprocessSVGs(spe)
dim(spe)
assayNames(spe)
```


# Identify SVGs

## HVGs (non-spatial)

Non-spatial HVGs for baseline comparison.

```{r}
# fit mean-variance relationship
dec <- modelGeneVar(spe)

# store in object
stopifnot(all(rownames(dec) == rowData(spe)$gene_id))
dec_keep <- dec[, c("bio", "p.value", "FDR")]
colnames(dec_keep) <- c("bio", "hvgs_pval", "hvgs_fdr")
rowData(spe) <- cbind(rowData(spe), dec_keep)
```

```{r}
# calculate ranks according to p-values
rowData(spe)$rank_hvgs_pval <- rank(dec$p.value, ties.method = "first")
# calculate ranks according to statistic value
rowData(spe)$rank_hvgs_bio <- rank(-1 * dec$bio, ties.method = "first")
```


## BRISC

Run SVGs method using BRISC (extensions version), and calculate likelihood ratio tests comparing models with and without spatial terms.

Runtime: ~4.5 hours on laptop with 4 cores

Note: either run code in first couple of code blocks below (long) runtime) or load pre-saved results object in next code block.

To do:

- add option for bootstrap significance tests

```{r}
# runtime_brisc <- system.time({
#   spe_brisc <- runSVGsBRISC(spe, x = NULL, n_threads = 10)
# })
```

```{r}
# runtime_brisc
```

```{r}
# file <- "../outputs/spe_brisc_spatzli.rds"
# saveRDS(spe_brisc, file = file)
```

```{r}
# alternatively: load previously saved results
spe_brisc <- readRDS("../outputs/spe_brisc_spatzli.rds")
```

```{r}
# calculate ranks according to p-values
rowData(spe_brisc)$rank_brisc_pval <- rank(rowData(spe_brisc)$pval, ties.method = "first")
# calculate ranks according to effect sizes
rowData(spe_brisc)$rank_brisc_sigmasq <- rank(-1 * rowData(spe_brisc)$sigma.sq, ties.method = "first")
rowData(spe_brisc)$rank_brisc_prop_sv <- rank(-1 * rowData(spe_brisc)$prop_sv, ties.method = "first")
```


## GpGp

Run SVGs method using GpGp, and calculate likelihood ratio tests comparing models with and without spatial terms.

Runtime: ~5 hours using 4 cores on laptop.

Note: either run code in first couple of code blocks below (5 hours runtime) or load pre-saved results object in next code block.

```{r}
# runtime_gpgp <- system.time({
#   spe_gpgp <- runSVGsGpGp(spe, x = NULL, n_threads = 10)
# })
```

```{r}
# runtime_gpgp
```

```{r}
# file <- "../outputs/spe_gpgp_spatzli.rds"
# saveRDS(spe_gpgp, file = file)
```

```{r}
# alternatively: load previously saved results
spe_gpgp <- readRDS("../outputs/spe_gpgp_spatzli.rds")
```

```{r}
# calculate ranks according to p-values
rowData(spe_gpgp)$rank_gpgp_pval <- rank(rowData(spe_gpgp)$pval, ties.method = "first")
# calculate ranks according to effect sizes
rowData(spe_gpgp)$rank_gpgp_sigmasq <- rank(-1 * rowData(spe_gpgp)$sigmasq, ties.method = "first")
rowData(spe_gpgp)$rank_gpgp_prop_sv <- rank(-1 * rowData(spe_gpgp)$prop_sv, ties.method = "first")
```


## SPARK-X

Note: not performing any additional filtering with SPARK functions

```{r}
# run SPARK-X
runtime_sparkx <- system.time({
  out_sparkx <- sparkx(
    counts(spe), 
    spatialCoords(spe), 
    numCores = 1, 
    option = "mixture"
  )
})

# SPARK-X runtime
runtime_sparkx

# SPARK-X results
head(out_sparkx$res_mtest)
```

```{r}
# store in SPE object
spe_sparkx <- spe
stopifnot(all(rownames(out_sparkx$res_mtest) == rowData(spe)$gene_id))
rowData(spe_sparkx) <- cbind(rowData(spe), out_sparkx$res_mtest)
```

```{r}
# calculate ranks according to p-values
rowData(spe_sparkx)$rank_sparkx_pval <- rank(rowData(spe_sparkx)$combinedPval, ties.method = "first")
```


---

# Comparisons

Comparisons between methods.

Notes:

- BRISC and GpGp are very similar
- BRISC/GpGp are more similar to HVGs than SPARK-X is similiar to HVGs

To do:

- Try running BRISC with different tolerance argument

```{r}
table(BRISC = rowData(spe_brisc)$pval == 0, 
      GpGp = rowData(spe_gpgp)$pval == 0)

table(BRISC = rowData(spe_brisc)$pval == 0, 
      HVGs = rowData(spe)$hvgs_pval <= 1e-12)

table(GpGp = rowData(spe_gpgp)$pval == 0, 
      HVGs = rowData(spe)$hvgs_pval <= 1e-6)

table(BRISC = rowData(spe_brisc)$pval == 0, 
      SparkX = rowData(spe_sparkx)$combinedPval <= 1e-12)

table(GpGp = rowData(spe_gpgp)$pval == 0, 
      SparkX = rowData(spe_sparkx)$combinedPval <= 1e-12)

table(SparkX = rowData(spe_sparkx)$combinedPval <= 1e-12, 
      HVGs = rowData(spe)$hvgs_pval <= 1e-12)
```


---

# Session info

```{r}
sessionInfo()
```

