---
title: "Comparison of SVGs methods: human DLPFC"
author: "Lukas Weber"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Introduction

Analyses for SVGs: comparison of methods including BRISC, GpGp, HVGs, using spatzli wrappers, human DLPFC dataset.

```{r, message = FALSE}
library(SpatialExperiment)
library(STexampleData)
library(spatzli)
library(scran)
library(SPARK)
library(dplyr)
library(ggplot2)
library(tidyr)
```


# Load data

```{r}
spe <- Visium_humanDLPFC()
dim(spe)
assayNames(spe)
```


# Preprocessing

Preprocessing steps: filter low-expressed genes, filter mitochondrial genes, normalization, and logcounts. Using convenience function in `spatzli` package.

```{r}
spe <- preprocessSVGs(spe)
dim(spe)
assayNames(spe)
```


# Identify SVGs

## HVGs (non-spatial)

Non-spatial HVGs for baseline comparison.

```{r}
# fit mean-variance relationship
dec <- modelGeneVar(spe)

# store in object
stopifnot(all(rownames(dec) == rowData(spe)$gene_id))
dec_keep <- dec[, c("bio", "p.value", "FDR")]
colnames(dec_keep) <- c("bio", "hvgs_pval", "hvgs_fdr")
rowData(spe) <- cbind(rowData(spe), dec_keep)
```

```{r}
# calculate ranks according to p-values
rowData(spe)$rank_hvgs_pval <- rank(dec$p.value, ties.method = "first")
# calculate ranks according to statistic value
rowData(spe)$rank_hvgs_bio <- rank(-1 * dec$bio, ties.method = "first")
```


## BRISC

To do:
- extract log-likelihoods from BRISC to enable likelihood ratio tests
- alternatively: use statistic values and effect estimates only


## GpGp

Run SVGs method using GpGp, and calculate likelihood ratio tests comparing models with and without spatial terms.

Runtime: ~5 hours using 4 cores on laptop.

Note: either run code in first couple of code blocks below (5 hours runtime) or load pre-saved results object in next code block.

```{r}
# runtime_gpgp <- system.time({
#   spe_gpgp <- runSVGsGpGp(spe, x = NULL, n_threads = 4)
# })
```

```{r}
# runtime_gpgp
```

```{r}
# file <- "../outputs/spe_gpgp_spatzli.rds"
# saveRDS(spe_gpgp, file = file)
```

```{r}
# alternatively: load previously saved results
spe_gpgp <- readRDS("../outputs/spe_gpgp_spatzli.rds")
```

```{r}
# calculate ranks according to p-values
rowData(spe_gpgp)$rank_gpgp_pval <- rank(rowData(spe_gpgp)$pval, ties.method = "first")
# calculate ranks according to effect sizes
rowData(spe_gpgp)$rank_gpgp_sigmasq <- rank(-1 * rowData(spe_gpgp)$sigmasq, ties.method = "first")
rowData(spe_gpgp)$rank_gpgp_prop_sv <- rank(-1 * rowData(spe_gpgp)$prop_sv, ties.method = "first")
```


## SPARK-X

Note: not performing any additional filtering with SPARK functions

```{r}
# run SPARK-X
runtime_sparkx <- system.time({
  out_sparkx <- sparkx(
    counts(spe), 
    spatialCoords(spe), 
    numCores = 1, 
    option = "mixture"
  )
})

# SPARK-X runtime
runtime_sparkx

# SPARK-X results
head(out_sparkx$res_mtest)
```

```{r}
# store in SPE object
spe_sparkx <- spe
stopifnot(all(rownames(out_sparkx$res_mtest) == rowData(spe)$gene_id))
rowData(spe_sparkx) <- cbind(rowData(spe), out_sparkx$res_mtest)
```

```{r}
# calculate ranks according to p-values
rowData(spe_sparkx)$rank_sparkx_pval <- rank(rowData(spe_sparkx)$combinedPval, ties.method = "first")
```


# Comparison

## Plots

### Set up data frame for plotting

```{r}
# data frame for plotting
stopifnot(all(rowData(spe)$gene_id == rowData(spe_gpgp)$gene_id))
stopifnot(all(rowData(spe)$gene_id == rowData(spe_sparkx)$gene_id))

dim(rowData(spe))
dim(rowData(spe_gpgp))
dim(rowData(spe_sparkx))

df <- as.data.frame(rowData(spe))
df <- full_join(df, as.data.frame(rowData(spe_gpgp)))
df <- full_join(df, as.data.frame(rowData(spe_sparkx)))

dim(df)
head(df, 3)
```

---

### GpGp statistic vs. HVGs statistic (values)

Using log(1 + x) transformation for improved visibility.

```{r, fig.width = 5, fig.height = 5}
ggplot(df, aes(x = bio, y = sigmasq)) + 
  geom_point(pch = 1) + 
  geom_smooth() + 
  scale_x_continuous(trans = "log1p") + 
  scale_y_continuous(trans = "log1p") + 
  labs(x = "HVGs statistic (bio)", 
       y = "GpGp statistic (sigmasq)") + 
  ggtitle("GpGp statistic vs. HVGs statistic") + 
  theme_bw()
```

---

### GpGp statistic vs. HVGs statistic (ranks)

**All genes**

```{r, fig.width = 5, fig.height = 5}
ggplot(df, aes(x = rank_hvgs_bio, y = rank_gpgp_sigmasq)) + 
  geom_point(pch = 1) + 
  geom_smooth() + 
  labs(x = "Rank of HVGs statistic (bio)", 
       y = "Rank of GpGp statistic (sigmasq)") + 
  ggtitle("Ranks GpGp statistic vs. rank HVGs statistic") + 
  theme_bw()
```

**Zoomed in to top 1000 genes from each method**

```{r, fig.width = 5, fig.height = 5}
ggplot(df, aes(x = rank_hvgs_bio, y = rank_gpgp_sigmasq)) + 
  geom_point(pch = 1) + 
  geom_smooth() + 
  lims(x = c(0, 1000), 
       y = c(0, 1000)) + 
  labs(x = "Rank of HVGs statistic (bio)", 
       y = "Rank of GpGp statistic (sigmasq)") + 
  ggtitle("Ranks GpGp statistic vs. rank HVGs statistic (top 1000)") + 
  theme_bw()
```

**Zoomed in to top 200 genes from each method**

```{r, fig.width = 5, fig.height = 5}
ggplot(df, aes(x = rank_hvgs_bio, y = rank_gpgp_sigmasq)) + 
  geom_point(pch = 1) + 
  geom_smooth() + 
  lims(x = c(0, 200), 
       y = c(0, 200)) + 
  labs(x = "Rank of HVGs statistic (bio)", 
       y = "Rank of GpGp statistic (sigmasq)") + 
  ggtitle("Ranks GpGp statistic vs. rank HVGs statistic (top 200)") + 
  theme_bw()
```

**Zoomed in to top 50 genes from each method**

```{r, fig.width = 5, fig.height = 5}
ggplot(df, aes(x = rank_hvgs_bio, y = rank_gpgp_sigmasq)) + 
  geom_point(pch = 1, size = 2, stroke = 1.5) + 
  lims(x = c(0, 50), 
       y = c(0, 50)) + 
  labs(x = "Rank of HVGs statistic (bio)", 
       y = "Rank of GpGp statistic (sigmasq)") + 
  ggtitle("Ranks GpGp statistic vs. rank HVGs statistic (top 50)") + 
  theme_bw()
```

Comments:
- the top genes from GpGp are highly overlapping with the top HVGs in this dataset (e.g top 50 genes)
- we see larger differences in the rankings at lower ranks (e.g. top 1000 genes)
- check differences vs. SPARK-X


---

### Significance (adjusted p-values) from SPARK-X vs. significance (adjusted p-values) from HVGs

**Zoomed in to adjusted p-values < 0.05**

```{r, fig.width = 5, fig.height = 5}
ggplot(df, aes(x = hvgs_fdr, y = adjustedPval)) + 
  geom_point(pch = 1) + 
  lims(x = c(0, 0.05), 
       y = c(0, 0.05)) + 
  labs(x = "HVGs: adjusted p-values", 
       y = "SPARK-X: adjusted p-values") + 
  ggtitle("Significance: SPARK-X vs. HVGs") + 
  theme_bw()
```

**Ranks: zoomed in to top 200 genes from each method**

```{r, fig.width = 5, fig.height = 5}
ggplot(df, aes(x = rank_hvgs_pval, y = rank_sparkx_pval)) + 
  geom_point(pch = 1) + 
  lims(x = c(0, 200), 
       y = c(0, 200)) + 
  labs(x = "HVGs: rank (p-values)", 
       y = "SPARK-X: rank (p-values)") + 
  ggtitle("Ranks: SPARK-X vs. HVGs") + 
  theme_bw()
```

**Ranks: zoomed in to top 200 genes from each method**

```{r, fig.width = 5, fig.height = 5}
ggplot(df, aes(x = rank_hvgs_bio, y = rank_sparkx_pval)) + 
  geom_point(pch = 1) + 
  lims(x = c(0, 200), 
       y = c(0, 200)) + 
  labs(x = "HVGs: rank (statistic)", 
       y = "SPARK-X: rank (p-values)") + 
  ggtitle("Ranks: SPARK-X vs. HVGs") + 
  theme_bw()
```

**Ranks: zoomed in to top 50 genes from each method**

```{r, fig.width = 5, fig.height = 5}
ggplot(df, aes(x = rank_hvgs_pval, y = rank_sparkx_pval)) + 
  geom_point(pch = 1) + 
  lims(x = c(0, 50), 
       y = c(0, 50)) + 
  labs(x = "HVGs: rank (p-values)", 
       y = "SPARK-X: rank (p-values)") + 
  ggtitle("Ranks: SPARK-X vs. HVGs") + 
  theme_bw()
```

**Ranks: zoomed in to top 50 genes from each method**

```{r, fig.width = 5, fig.height = 5}
ggplot(df, aes(x = rank_hvgs_bio, y = rank_sparkx_pval)) + 
  geom_point(pch = 1) + 
  lims(x = c(0, 50), 
       y = c(0, 50)) + 
  labs(x = "HVGs: rank (statistic)", 
       y = "SPARK-X: rank (p-values)") + 
  ggtitle("Ranks: SPARK-X vs. HVGs") + 
  theme_bw()
```

---

### Significance (adjusted p-values) from SPARK-X vs. significance (adjusted p-values) from GpGp

**Zoomed in to adjusted p-values < 0.05**

```{r, fig.width = 5, fig.height = 5}
ggplot(df, aes(x = padj, y = adjustedPval)) + 
  geom_point(pch = 1) + 
  lims(x = c(0, 0.05), 
       y = c(0, 0.05)) + 
  labs(x = "GpGp: adjusted p-values", 
       y = "SPARK-X: adjusted p-values") + 
  ggtitle("Significance: SPARK-X vs. GpGp") + 
  theme_bw()
```

**Ranks: Zoomed in to top 200 genes**

```{r, fig.width = 5, fig.height = 5}
ggplot(df, aes(x = rank_gpgp_pval, y = rank_sparkx_pval)) + 
  geom_point(pch = 1) + 
  lims(x = c(0, 200), 
       y = c(0, 200)) + 
  labs(x = "GpGp: ranks (p-values)", 
       y = "SPARK-X: ranks (p-values)") + 
  ggtitle("Ranks: SPARK-X vs. GpGp") + 
  theme_bw()
```


---

### Significance (adjusted p-values) vs. effect size (proportion spatial variance) for GpGp method

**All genes**

```{r, fig.width = 5, fig.height = 5}
ggplot(df, aes(x = prop_sv, y = padj)) + 
  geom_point(pch = 1) + 
  labs(x = "proportion spatial variance", 
       y = "adjusted p-value") + 
  ggtitle("Significance vs. effect size: GpGp method") + 
  theme_bw()
```

**Top-ranked genes from GpGp (according to significance)**

```{r, fig.width = 5, fig.height = 5}
ix_keep <- df$rank_gpgp_pval <= 500
table(ix_keep)

ggplot(df[ix_keep, ], aes(x = prop_sv, y = padj)) + 
  geom_point(pch = 1) + 
  labs(x = "proportion spatial variance", 
       y = "adjusted p-value") + 
  ggtitle("Significance vs. effect size: GpGp method") + 
  theme_bw()
```

**Top-ranked genes from GpGp (according to statistic)**

```{r, fig.width = 5, fig.height = 5}
ix_keep <- df$rank_gpgp_sigmasq <= 500
table(ix_keep)

ggplot(df[ix_keep, ], aes(x = prop_sv, y = padj)) + 
  geom_point(pch = 1) + 
  labs(x = "proportion spatial variance", 
       y = "adjusted p-value") + 
  ggtitle("Significance vs. effect size: GpGp method") + 
  theme_bw()
```

Comments:
- top-ranked genes from GpGp all have p-value == 0, so cannot distinguish these easily using GpGp; may be possible with BRISC instead


---

### Histogram of p-values

```{r, fig.width = 5, fig.height = 4}
ggplot(df, aes(x = padj)) + 
  geom_histogram(bins = 40, color = "black", fill = "darkorange2") + 
  labs(x = "adjusted p-values") + 
  ggtitle("Histogram of adjusted p-values") + 
  theme_bw()
```

Comments:
- large numbers of genes with p-values == 0 or 1 with GpGp method


---

### Effect sizes (proportion spatial variance) for genes with p-values exactly equal to zero for GpGp method

```{r, fig.width = 5, fig.height = 4}
# number of genes with p-value == 0
table(df$padj == 0)

df_sub <- df[df$padj == 0, ]
dim(df_sub)

ggplot(df_sub, aes(x = prop_sv)) + 
  geom_histogram(bins = 40, color = "black", fill = "navy") + 
  geom_vline(xintercept = 0.2, color = "red") + 
  labs(x = "proportion spatial variance") + 
  ggtitle("Effect sizes for genes with p-values == 0: GpGp") + 
  theme_bw()
```


## Significance rule

**Example of overall significance rule using both statistical significance and effect size. Six known interesting genes in this dataset (SNAP25, PCP4, MOBP, HBB, IGKC, NPY) are all detected by this rule.**

```{r, fig.width = 5, fig.height = 4}
# number of genes with p-value == 0
table(df$padj == 0)
# number of genes with effect size >= 0.2
table(df$prop_sv >= 0.2)

# number of genes with 'p-value == 0' & 'effect size >= 0.2'
table(df$padj == 0 & df$prop_sv >= 0.2)
# contingency table
table(df$padj == 0, df$prop_sv >= 0.2)

# print gene names
sig <- df$padj == 0 & df$prop_sv >= 0.2
sig_names <- rowData(spe)$gene_name[sig]
sig_names_alph <- sort(sig_names)
sig_names_alph

# check known interesting genes in this dataset
interesting <- c("SNAP25", "PCP4", "MOBP", "HBB", "IGKC", "NPY")
interesting %in% sig_names
```


## Plot significant genes

### Plot expression for all genes that meet the overall significance rule


**Ordered by effect size**

```{r, fig.width = 2.5, fig.height = 2.5}
stopifnot(all(df$gene_name == rowData(spe)$gene_name))
stopifnot(all(df$gene_name == rowData(spe_gpgp)$gene_name))

df_sub <- df[sig, ]
ix_sort <- order(df[sig, ]$prop_sv, decreasing = TRUE)
df_sub <- df_sub[ix_sort, ]
head(df_sub)

for (i in seq_len(nrow(df_sub))) {
  gene <- df_sub$gene_name[i]
  ix_gene <- which(rowData(spe)$gene_name == gene)
  df_spots <- as.data.frame(cbind(
    spatialCoords(spe), 
    logexpr = logcounts(spe)[ix_gene, ]
  ))
  
  p <- ggplot(df_spots, aes(x = x, y = y, color = logexpr)) + 
    geom_point(size = 0.1) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_gradient(low = "gray95", high = "blue") + 
    ggtitle(gene) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


---

### Plot expression of top genes from SPARK-X for comparison

```{r, fig.width = 2.5, fig.height = 2.5}
stopifnot(all(df$gene_name == rowData(spe)$gene_name))
stopifnot(all(df$gene_name == rowData(spe_sparkx)$gene_name))

ix_keep <- which(rowData(spe_sparkx)$rank_sparkx_pval %in% 1:100)
df_sub <- df[ix_keep, ]
ix_sort <- order(df_sub$rank_sparkx_pval)
df_sub <- df_sub[ix_sort, ]
head(df_sub, 3)
top_genes_sparkx <- df_sub$gene_name

# top genes (in order)
print(top_genes_sparkx)

for (gene in top_genes_sparkx) {
  ix_gene <- which(rowData(spe)$gene_name == gene)
  df_spots <- as.data.frame(cbind(
    spatialCoords(spe), 
    logexpr = logcounts(spe)[ix_gene, ]
  ))
  
  p <- ggplot(df_spots, aes(x = x, y = y, color = logexpr)) + 
    geom_point(size = 0.1) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_gradient(low = "gray95", high = "purple4") + 
    ggtitle(gene) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


---

### Plot expression of top HVGs for comparison


**Top ranks by p-value**

```{r, fig.width = 2.5, fig.height = 2.5}
stopifnot(all(df$gene_name == rowData(spe)$gene_name))

ix_keep <- which(rowData(spe)$rank_hvgs_pval %in% 1:100)
df_sub <- df[ix_keep, ]
ix_sort <- order(df_sub$rank_hvgs_pval)
df_sub <- df_sub[ix_sort, ]
head(df_sub, 3)
top_genes_hvgs <- df_sub$gene_name

# top genes (in order)
print(top_genes_hvgs)

for (gene in top_genes_hvgs) {
  ix_gene <- which(rowData(spe)$gene_name == gene)
  df_spots <- as.data.frame(cbind(
    spatialCoords(spe), 
    logexpr = logcounts(spe)[ix_gene, ]
  ))
  
  p <- ggplot(df_spots, aes(x = x, y = y, color = logexpr)) + 
    geom_point(size = 0.1) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_gradient(low = "gray95", high = "forestgreen") + 
    ggtitle(gene) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


**Top ranks by statistic value**

```{r, fig.width = 2.5, fig.height = 2.5}
stopifnot(all(df$gene_name == rowData(spe)$gene_name))

ix_keep <- which(rowData(spe)$rank_hvgs_bio %in% 1:100)
df_sub <- df[ix_keep, ]
ix_sort <- order(df_sub$rank_hvgs_bio)
df_sub <- df_sub[ix_sort, ]
head(df_sub, 3)
top_genes_hvgs <- df_sub$gene_name

# top genes (in order)
print(top_genes_hvgs)

for (gene in top_genes_hvgs) {
  ix_gene <- which(rowData(spe)$gene_name == gene)
  df_spots <- as.data.frame(cbind(
    spatialCoords(spe), 
    logexpr = logcounts(spe)[ix_gene, ]
  ))
  
  p <- ggplot(df_spots, aes(x = x, y = y, color = logexpr)) + 
    geom_point(size = 0.1) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_gradient(low = "gray95", high = "forestgreen") + 
    ggtitle(gene) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


---

## Gene clustering

### GpGp genes: k-means

**Clustering genes from GpGp with adjusted p-values <= 0.05**

```{r}
# number of genes with adjusted p-value == 0
table(df$padj == 0)
# number of genes with adjusted p-value <= 0.05
table(df$padj <= 0.05)

stopifnot(all(df$gene_id == rowData(spe)$gene_id))
ix_sig_gpgp <- df$padj <= 0.05

# subset genes
spe_sig_gpgp <- spe[ix_sig_gpgp, ]
dim(spe_sig_gpgp)

# z-score normalize logcounts from these genes
znorm <- t(apply(logcounts(spe_sig_gpgp), 1, scale, center = TRUE, scale = TRUE))
dim(znorm)
znorm[1:3, 1:3]
```


```{r, fig.width = 2.5, fig.height = 2.5}
# cluster on genes
k <- 10
kmeans_out <- kmeans(znorm, centers = k)
clus <- kmeans_out$cluster
length(clus)
table(clus)

# cluster centers
centers <- kmeans_out$centers
dim(centers)
centers[1:3, 1:3]

# plot cluster centers
df_centers <- t(centers)
rownames(df_centers) <- colnames(spe_sig_gpgp)
colnames(df_centers) <- paste0("profile", seq_len(k))
df_centers <- cbind(df_centers, spatialCoords(spe_sig_gpgp))
df_centers <- as.data.frame(df_centers)
df_centers <- gather(df_centers, key = "profile", value = "znorm", paste0("profile", seq_len(k)))
df_centers$profile <- as.factor(df_centers$profile)
head(df_centers)
str(df_centers)

for (g in seq_len(k)) {
  df_centers_sub <- df_centers[df_centers$profile == paste0("profile", g), ]
  
  p <- ggplot(df_centers_sub, aes(x = x, y = y, color = znorm)) + 
    geom_point(size = 0.1) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_gradient2(low = "blue", mid = "white", high = "red") + 
    ggtitle(paste0("GpGp profile", g)) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


```{r, fig.width = 2.5, fig.height = 2.5}
# cluster on genes
k <- 50
kmeans_out <- kmeans(znorm, centers = k)
clus <- kmeans_out$cluster
length(clus)
table(clus)

# cluster centers
centers <- kmeans_out$centers
dim(centers)
centers[1:3, 1:3]

# plot cluster centers
df_centers <- t(centers)
rownames(df_centers) <- colnames(spe_sig_gpgp)
colnames(df_centers) <- paste0("profile", seq_len(k))
df_centers <- cbind(df_centers, spatialCoords(spe_sig_gpgp))
df_centers <- as.data.frame(df_centers)
df_centers <- gather(df_centers, key = "profile", value = "znorm", paste0("profile", seq_len(k)))
df_centers$profile <- as.factor(df_centers$profile)
head(df_centers)
str(df_centers)

for (g in seq_len(k)) {
  df_centers_sub <- df_centers[df_centers$profile == paste0("profile", g), ]
  
  p <- ggplot(df_centers_sub, aes(x = x, y = y, color = znorm)) + 
    geom_point(size = 0.1) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_gradient2(low = "blue", mid = "white", high = "red") + 
    ggtitle(paste0("GpGp profile", g)) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


### GpGp genes: hierarchical

**Clustering genes from GpGp with adjusted p-values == 0**

Since hierarchical clustering requires storing distance matrix in memory, we use a smaller number of genes.

```{r}
# subset genes
spe_sig_gpgp <- spe[df$padj == 0, ]

dim(spe_sig_gpgp)

# z-score normalize logcounts from these genes
znorm <- t(apply(logcounts(spe_sig_gpgp), 1, scale, center = TRUE, scale = TRUE))
dim(znorm)
znorm[1:3, 1:3]
```

```{r}
# cluster on genes
d <- dist(t(znorm), method = "euclidean")
str(d)

# hierarchical clustering using Ward's linkage
hc <- hclust(d, method = "ward.D2")

# select 12 clusters
k <- 12
ct <- cutree(hc, k = 12)

table(ct)
```


---

### SPARK-X genes

**Clustering genes from SPARK-X with adjusted p-values <= 0.05**

```{r}
# number of genes with adjusted p-value == 0
table(df$adjustedPval == 0)
# number of genes with adjusted p-value <= 0.05
table(df$adjustedPval <= 0.05)

stopifnot(all(df$gene_id == rowData(spe)$gene_id))
ix_sig_sparkx <- df$adjustedPval <= 0.05

# subset genes
spe_sig_sparkx <- spe[ix_sig_sparkx, ]
dim(spe_sig_sparkx)

# z-score normalize logcounts from these genes
znorm <- t(apply(logcounts(spe_sig_sparkx), 1, scale, center = TRUE, scale = TRUE))
dim(znorm)
znorm[1:3, 1:3]
```


```{r, fig.width = 2.5, fig.height = 2.5}
# cluster on genes
k <- 10
kmeans_out <- kmeans(znorm, centers = k)
clus <- kmeans_out$cluster
length(clus)
table(clus)

# cluster centers
centers <- kmeans_out$centers
dim(centers)
centers[1:3, 1:3]

# plot cluster centers
df_centers <- t(centers)
rownames(df_centers) <- colnames(spe_sig_sparkx)
colnames(df_centers) <- paste0("profile", seq_len(k))
df_centers <- cbind(df_centers, spatialCoords(spe_sig_sparkx))
df_centers <- as.data.frame(df_centers)
df_centers <- gather(df_centers, key = "profile", value = "znorm", paste0("profile", seq_len(k)))
df_centers$profile <- as.factor(df_centers$profile)
head(df_centers)
str(df_centers)

for (g in seq_len(k)) {
  df_centers_sub <- df_centers[df_centers$profile == paste0("profile", g), ]
  
  p <- ggplot(df_centers_sub, aes(x = x, y = y, color = znorm)) + 
    geom_point(size = 0.1) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_gradient2(low = "blue", mid = "white", high = "red") + 
    ggtitle(paste0("SPARK-X profile", g)) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


```{r, fig.width = 2.5, fig.height = 2.5}
# cluster on genes
k <- 50
kmeans_out <- kmeans(znorm, centers = k)
clus <- kmeans_out$cluster
length(clus)
table(clus)

# cluster centers
centers <- kmeans_out$centers
dim(centers)
centers[1:3, 1:3]

# plot cluster centers
df_centers <- t(centers)
rownames(df_centers) <- colnames(spe_sig_sparkx)
colnames(df_centers) <- paste0("profile", seq_len(k))
df_centers <- cbind(df_centers, spatialCoords(spe_sig_sparkx))
df_centers <- as.data.frame(df_centers)
df_centers <- gather(df_centers, key = "profile", value = "znorm", paste0("profile", seq_len(k)))
df_centers$profile <- as.factor(df_centers$profile)
head(df_centers)
str(df_centers)

for (g in seq_len(k)) {
  df_centers_sub <- df_centers[df_centers$profile == paste0("profile", g), ]
  
  p <- ggplot(df_centers_sub, aes(x = x, y = y, color = znorm)) + 
    geom_point(size = 0.1) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_gradient2(low = "blue", mid = "white", high = "red") + 
    ggtitle(paste0("SPARK-X profile", g)) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


---

# Session info

```{r}
sessionInfo()
```

