---
title: "SVGs BRISC analyses: mouse olfactory"
author: "Lukas Weber"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Introduction

Analyses for SVGs BRISC method: mouse olfactory bulb dataset

Note: installation of MERINGUE and SPARK both require compilation

```{r, message=FALSE}
library(SpatialExperiment)
library(dplyr)
library(ggplot2)
```


## Load data

Load dataset from MERINGUE package and reshape into `SpatialExperiment` format.

```{r, message=FALSE}
library(MERINGUE)
```

```{r}
# load dataset from MERINGUE package
data(mOB)

names(mOB)
dim(mOB$counts)
mOB$counts[1000:1006, 1:6]
head(mOB$pos)
head(mOB$results)
str(mOB$annot)

# create SpatialExperiment
row_data = data.frame(
  gene_name = rownames(mOB$counts)
)
res <- cbind(mOB$results, gene_name = rownames(mOB$results))
row_data <- DataFrame(left_join(row_data, res, by = "gene_name"))
head(row_data)

col_data <- data.frame(
  barcode_id = colnames(mOB$counts)
)
annot <- data.frame(
  barcode_id = names(mOB$annot), 
  layer_id = mOB$annot
)
col_data <- DataFrame(left_join(col_data, annot, by = "barcode_id"))
head(col_data)

spatial_coords <- as.matrix(mOB$pos[col_data$barcode_id, ])
head(spatial_coords)

stopifnot(nrow(spatial_coords) == nrow(col_data))
stopifnot(all(rownames(spatial_coords) == col_data$barcode_id))
stopifnot(all(colnames(mOB$counts) == col_data$barcode_id))

spe <- SpatialExperiment(
  assays = list(counts = mOB$counts), 
  rowData = row_data, 
  colData = col_data, 
  spatialCoords = spatial_coords
)

spe
```


## Preprocessing

Re-using code from OSTA:

- keep all genes, not just HVGs

```{r, message=FALSE}
# QUALITY CONTROL (QC)

library(scater)
# mitochondrial genes: none included in this dataset
is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name)
table(is_mito)
# spot-level filtering
spe <- addPerCellQC(spe)
qc_lib_size <- colData(spe)$sum < 100
qc_detected <- colData(spe)$detected < 100
discard <- qc_lib_size | qc_detected
table(discard)
colData(spe)$discard <- discard
# check
colData(spe)[colData(spe)$discard, ]
# filter low-quality spots
spe <- spe[, !colData(spe)$discard]
dim(spe)

# NORMALIZATION

library(scran)
# calculate size factors
spe <- computeSumFactors(spe)
# calculate logcounts (log-transformed normalized counts)
spe <- logNormCounts(spe)

# FEATURE SELECTION

# fit mean-variance relationship
dec <- modelGeneVar(spe)
# select top HVGs
top_hvgs <- getTopHVGs(dec, prop = 0.1)
str(top_hvgs)
# keep HVG statistics
rowData(spe) <- DataFrame(left_join(
  as.data.frame(rowData(spe)), as.data.frame(cbind(dec, gene_name = rownames(dec))), 
  by = "gene_name"
))
head(rowData(spe))
```


## Run SVGs methods

### Run BRISC

Using parallelized wrapper function in `spatzli` package.

notes:
- runtime for bootstrapping to get confidence intervals is ~100x slower
- to do: add slower version to rank SVGs by significance of `sigma.sq` parameter (using bootstrap confidence intervals)

runtime: ~7 min on laptop (4 threads)

```{r}
library(spatzli)
```

```{r}
# run BRISC to calculate statistics for each gene
runtime_brisc <- system.time({
  spe <- rankSVGsBRISC(spe, n_threads = 4)
})

runtime_brisc
```


### Run BRISC with covariates

Runtime: ~7 min (4 threads) on laptop

```{r}
# run BRISC to calculate statistics for each gene
runtime_brisc_covariates <- system.time({
  spe_covariates <- rankSVGsBRISC(spe, x = model.matrix(~ colData(spe)$layer_id), n_threads = 4)
})

runtime_brisc_covariates
```


### Run SPARK-X

```{r, message=FALSE}
library(SPARK)
```

```{r}
runtime_sparkx <- system.time({
  sparkx_out <- sparkx(
    counts(spe), spatialCoords(spe), numCores = 4, option="mixture"
  )
})

runtime_sparkx

names(sparkx_out)

stopifnot(all(rownames(sparkx_out$stats) == rowData(spe)$gene_name))

rowData(spe) <- cbind(DataFrame(
  rowData(spe), sparkx_out$stats, sparkx_out$res_stest, sparkx_out$res_mtest
))
```


### Run MERINGUE

Use stored results in object provided with MERINGUE package.


### Calculate statistic ranks

Note: SPARK-X does not return a final combined statistic

```{r}
# calculate ranks

# BRISC
rowData(spe)$rank_svgs_sigmasq <- rank(-1 * rowData(spe)$sigma.sq, na.last = "keep")

# BRISC with covariates
all(rownames(spe) == rownames(spe_covariates))
rowData(spe)$rank_svgs_sigmasq_covariates <- rank(-1 * rowData(spe_covariates)$sigma.sq, na.last = "keep")

# MERINGUE
rowData(spe)$rank_svgs_I <- rank(-1 * rowData(spe)$I, na.last = "keep")

# HVGs
rowData(spe)$rank_hvgs_bio <- rank(-1 * rowData(spe)$bio, na.last = "keep")

# SPARK-X (using p-values instead of statistic)
rowData(spe)$rank_svgs_sparkx <- rank(rowData(spe)$combinedPval, na.last = "keep")
```


## Comparisons

### BRISC vs. MERINGUE

```{r, fig.width=4.25, fig.height=4}
ggplot(as.data.frame(rowData(spe)), 
       aes(x = I, y = sigma.sq)) + 
  geom_point(pch = 1) + 
  geom_smooth() + 
  labs(x = "MERINGUE: Moran's I", 
       y = "BRISC: sigma.sq") + 
  ggtitle("Statistics") + 
  theme_bw()

ggplot(as.data.frame(rowData(spe)), 
       aes(x = rank_svgs_I, y = rank_svgs_sigmasq)) + 
  geom_point(pch = 1) + 
  geom_smooth() + 
  labs(x = "MERINGUE: Moran's I (ranks)", 
       y = "BRISC: sigma.sq (ranks)") + 
  ggtitle("Ranks") + 
  theme_bw()
```


### BRISC vs. non-spatial HVGs

```{r, fig.width=4.25, fig.height=4}
ggplot(as.data.frame(rowData(spe)), 
       aes(x = bio, y = sigma.sq)) + 
  geom_point(pch = 1) + 
  geom_smooth() + 
  labs(x = "HVGs: bio", 
       y = "BRISC: sigma.sq") + 
  ggtitle("Statistics") + 
  theme_bw()

ggplot(as.data.frame(rowData(spe)), 
       aes(x = rank_hvgs_bio, y = rank_svgs_sigmasq)) + 
  geom_point(pch = 1) + 
  geom_smooth() + 
  labs(x = "HVGs: bio (ranks)", 
       y = "BRISC: sigma.sq (ranks)") + 
  ggtitle("Ranks") + 
  theme_bw()
```


### MERINGUE vs. non-spatial HVGs

```{r, fig.width=4.25, fig.height=4}
ggplot(as.data.frame(rowData(spe)), 
       aes(x = bio, y = I)) + 
  geom_point(pch = 1) + 
  geom_smooth() + 
  labs(x = "HVGs: bio", 
       y = "MERINGUE: Moran's I") + 
  ggtitle("Statistics") + 
  theme_bw()

ggplot(as.data.frame(rowData(spe)), 
       aes(x = rank_hvgs_bio, y = rank_svgs_I)) + 
  geom_point(pch = 1) + 
  geom_smooth() + 
  labs(x = "HVGs: bio (ranks)", 
       y = "MERINGUE: Moran's I (ranks)") + 
  ggtitle("Ranks") + 
  theme_bw()
```


### SPARK-X vs. MERINGUE (p-values)

```{r, fig.width=4.25, fig.height=4}
ggplot(as.data.frame(rowData(spe)), 
       aes(x = Adjusted.P.value, y = adjustedPval)) + 
  geom_point(pch = 1) + 
  geom_smooth() + 
  labs(x = "MERINGUE: adjusted p-values", 
       y = "SPARK-X: adjusted p-values") + 
  ggtitle("P-values") + 
  theme_bw()
```


## Plots

### Top genes for each method

Top genes differ between methods

```{r}
# identify top gene for each method
ix_top_brisc <- which.max(rowData(spe)$sigma.sq)
ix_top_meringue <- which.max(rowData(spe)$I)
ix_top_hvgs <- which.max(rowData(spe)$bio)
ix_top_sparkx <- which.min(rowData(spe)$combinedPval)

ix_top_brisc_covariates <- which.max(rowData(spe_covariates)$sigma.sq)

rowData(spe)$gene_name[ix_top_brisc]
rowData(spe)$gene_name[ix_top_meringue]
rowData(spe)$gene_name[ix_top_hvgs]
rowData(spe)$gene_name[ix_top_sparkx]

rowData(spe_covariates)$gene_name[ix_top_brisc_covariates]
```


Top 10 genes from each method

```{r}
# identify top gene for each method
ix_top10_brisc <- which(rowData(spe)$rank_svgs_sigmasq <= 10)
ix_top10_meringue <- which(rowData(spe)$rank_svgs_I <= 10)
ix_top10_hvgs <- which(rowData(spe)$rank_hvgs_bio <= 10)
ix_top10_sparkx <- which(rowData(spe)$rank_svgs_sparkx <= 10)

rowData(spe)$gene_name[ix_top10_brisc]
rowData(spe)$gene_name[ix_top10_meringue]
rowData(spe)$gene_name[ix_top10_hvgs]
rowData(spe)$gene_name[ix_top10_sparkx]
```


### BRISC

```{r, fig.width=5, fig.height=4}
ggplot(as.data.frame(cbind(spatialCoords(spe), counts = counts(spe)[ix_top_brisc, ])), 
       aes(x = x, y = y, color = counts)) + 
  coord_fixed() + 
  geom_point(pch = 16, size = 4.5) + 
  scale_color_gradient(low = "gray90", high = "blue") + 
  ggtitle(paste0("BRISC top gene: ", rowData(spe)$gene_name[ix_top_brisc])) + 
  theme_bw()
```


### MERINGUE

```{r, fig.width=5, fig.height=4}
ggplot(as.data.frame(cbind(spatialCoords(spe), counts = counts(spe)[ix_top_meringue, ])), 
       aes(x = x, y = y, color = counts)) + 
  coord_fixed() + 
  geom_point(pch = 16, size = 4.5) + 
  scale_color_gradient(low = "gray90", high = "blue") + 
  ggtitle(paste0("MERINGUE top gene: ", rowData(spe)$gene_name[ix_top_meringue])) + 
  theme_bw()
```


### Non-spatial HVGs

```{r, fig.width=5, fig.height=4}
ggplot(as.data.frame(cbind(spatialCoords(spe), counts = counts(spe)[ix_top_hvgs, ])), 
       aes(x = x, y = y, color = counts)) + 
  coord_fixed() + 
  geom_point(pch = 16, size = 4.5) + 
  scale_color_gradient(low = "gray90", high = "blue") + 
  ggtitle(paste0("HVGs top gene: ", rowData(spe)$gene_name[ix_top_hvgs])) + 
  theme_bw()
```


### SPARK-X

```{r, fig.width=5, fig.height=4}
ggplot(as.data.frame(cbind(spatialCoords(spe), counts = counts(spe)[ix_top_sparkx, ])), 
       aes(x = x, y = y, color = counts)) + 
  coord_fixed() + 
  geom_point(pch = 16, size = 4.5) + 
  scale_color_gradient(low = "gray90", high = "blue") + 
  ggtitle(paste0("SPARK-X top gene: ", rowData(spe)$gene_name[ix_top_sparkx])) + 
  theme_bw()
```


### BRISC with covariates

```{r, fig.width=5, fig.height=4}
ggplot(as.data.frame(cbind(spatialCoords(spe_covariates), counts = counts(spe_covariates)[ix_top_brisc_covariates, ])), 
       aes(x = x, y = y, color = counts)) + 
  coord_fixed() + 
  geom_point(pch = 16, size = 4.5) + 
  scale_color_gradient(low = "gray90", high = "blue") + 
  ggtitle(paste0("BRISC (with covariates): ", rowData(spe_covariates)$gene_name[ix_top_brisc_covariates])) + 
  theme_bw()
```


## Overlapping genes

Plot number of overlapping genes for increasing number of genes.

### Overlap with HVGs as baseline

```{r}
pal_okabeito <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r}
df_overlaps_HVGs <- data.frame(
  HVGs = c(10, 50, 100, 200, 500), 
  BRISC = c(
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 10, ]$rank_svgs_sigmasq <= 10), 
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 50, ]$rank_svgs_sigmasq <= 50), 
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 100, ]$rank_svgs_sigmasq <= 100), 
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 200, ]$rank_svgs_sigmasq <= 200), 
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 500, ]$rank_svgs_sigmasq <= 500)
  ), 
  BRISC_covariates = c(
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 10, ]$rank_svgs_sigmasq_covariates <= 10), 
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 50, ]$rank_svgs_sigmasq_covariates <= 50), 
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 100, ]$rank_svgs_sigmasq_covariates <= 100), 
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 200, ]$rank_svgs_sigmasq_covariates <= 200), 
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 500, ]$rank_svgs_sigmasq_covariates <= 500)
  ), 
  moransI = c(
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 10, ]$rank_svgs_I <= 10, na.rm = TRUE), 
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 50, ]$rank_svgs_I <= 50, na.rm = TRUE), 
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 100, ]$rank_svgs_I <= 100, na.rm = TRUE), 
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 200, ]$rank_svgs_I <= 200, na.rm = TRUE), 
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 500, ]$rank_svgs_I <= 500, na.rm = TRUE)
  ), 
  sparkX = c(
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 10, ]$rank_svgs_sparkx <= 10), 
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 50, ]$rank_svgs_sparkx <= 50), 
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 100, ]$rank_svgs_sparkx <= 100), 
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 200, ]$rank_svgs_sparkx <= 200), 
    sum(rowData(spe)[rowData(spe)$rank_hvgs_bio <= 500, ]$rank_svgs_sparkx <= 500)
  )
)
```


```{r, fig.width=6, fig.height=4}
library(tidyr)

df_overlaps_HVGs <- 
  df_overlaps_HVGs %>% 
  gather("HVGs", "overlap")
colnames(df_overlaps_HVGs)[1] <- "method"
df_overlaps_HVGs$method <- as.factor(df_overlaps_HVGs$method)
df_overlaps_HVGs$size <- rep(c(10, 50, 100, 200, 500), times = 4)

ggplot(df_overlaps_HVGs, 
       aes(x = size, y = overlap, color = method)) + 
  geom_line() + 
  geom_point() + 
  scale_color_manual(values = pal_okabeito) + 
  xlab("Top k genes (HVGs)") + 
  ylab("Overlap size") + 
  ggtitle("Overlap with HVGs") + 
  theme_bw()

ggsave("../plots/overlaps_HVGs.pdf", width = 6, height = 4)
```


### Overlap with SPARK-X as baseline


```{r}
df_overlaps_sparkX <- data.frame(
  sparkX = c(10, 50, 100, 200, 500), 
  BRISC = c(
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 10, ]$rank_svgs_sigmasq <= 10), 
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 50, ]$rank_svgs_sigmasq <= 50), 
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 100, ]$rank_svgs_sigmasq <= 100), 
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 200, ]$rank_svgs_sigmasq <= 200), 
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 500, ]$rank_svgs_sigmasq <= 500)
  ), 
  BRISC_covariates = c(
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 10, ]$rank_svgs_sigmasq_covariates <= 10), 
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 50, ]$rank_svgs_sigmasq_covariates <= 50), 
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 100, ]$rank_svgs_sigmasq_covariates <= 100), 
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 200, ]$rank_svgs_sigmasq_covariates <= 200), 
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 500, ]$rank_svgs_sigmasq_covariates <= 500)
  ), 
  moransI = c(
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 10, ]$rank_svgs_I <= 10, na.rm = TRUE), 
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 50, ]$rank_svgs_I <= 50, na.rm = TRUE), 
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 100, ]$rank_svgs_I <= 100, na.rm = TRUE), 
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 200, ]$rank_svgs_I <= 200, na.rm = TRUE), 
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 500, ]$rank_svgs_I <= 500, na.rm = TRUE)
  ), 
  HVGs = c(
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 10, ]$rank_hvgs_bio <= 10), 
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 50, ]$rank_hvgs_bio <= 50), 
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 100, ]$rank_hvgs_bio <= 100), 
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 200, ]$rank_hvgs_bio <= 200), 
    sum(rowData(spe)[rowData(spe)$rank_svgs_sparkx <= 500, ]$rank_hvgs_bio <= 500)
  )
)
```


```{r, fig.width=6, fig.height=4}
library(tidyr)

df_overlaps_sparkX <- 
  df_overlaps_sparkX %>% 
  gather("sparkX", "overlap")
colnames(df_overlaps_sparkX)[1] <- "method"
df_overlaps_sparkX$method <- as.factor(df_overlaps_sparkX$method)
df_overlaps_sparkX$size <- rep(c(10, 50, 100, 200, 500), times = 4)

ggplot(df_overlaps_sparkX, 
       aes(x = size, y = overlap, color = method)) + 
  geom_line() + 
  geom_point() + 
  scale_color_manual(values = pal_okabeito[c(1, 2, 8, 4)]) + 
  xlab("Top k genes (SPARK-X)") + 
  ylab("Overlap size") + 
  ggtitle("Overlap with SPARK-X") + 
  theme_bw()

ggsave("../plots/overlaps_sparkX.pdf", width = 6, height = 4)
```


## To do

- SPARK-X: investigate individual statistics instead of combined p-values
- BRISC: improve filtering - remove low-expressed genes to speed up runtime
- BRISC with covariates: check code in spatzli package
- scalability: compare scalability - BRISC linear in number of spots? SPARK-X fastest for smaller datasets?
- DLPFC dataset (with / without covariates for layer IDs)


## Session info

```{r}
sessionInfo()
```

