---
title: "SVGs BRISC analyses"
author: "Lukas Weber"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

Analyses for SVGs BRISC method - including evaluations of test statistic properties, comparisons with HVGs, and runtimes.

```{r, eval=FALSE}
# start session on cluster
screen -S svgs_analyses
qrsh -l mem_free=10G,h_vmem=12G,h_fsize=100G
cd /dcs04/hicks/data/lweber/SVGs_spNNGP/SVGs-spNNGP-analyses
module load conda_R/devel
R
```


## Load results

Load results saved in object from script `SVGs_BRISC_humanDLPFC_spatzli.R`. The results are stored in the `rowData` of the `SpatialExperiment` object.

```{r, message=FALSE}
library(SpatialExperiment)
library(spatzli)

spe <- readRDS("../outputs/spe_brisc.rds")
```


## Evaluations of statistic

Evaluations of the properties of the statistic across genes.

```{r, message=FALSE}
library(ggplot2)
library(ggrepel)
```

```{r}
# calculate total UMIs per gene
rowData(spe)$sum <- rowSums(counts(spe))
# calculate mean logcounts per gene
rowData(spe)$mean <- rowMeans(logcounts(spe))

# loess fit to mean logcounts
# note: handle NAs
is_nas <- is.na(rowData(spe)$sigma.sq)
trend <- predict(loess(rowData(spe)$sigma.sq[!is_nas] ~ rowData(spe)$mean[!is_nas]))
rowData(spe)$trend <- NA
rowData(spe)$trend[!is_nas] <- trend

# calculate difference above trend
rowData(spe)$difftrend <- rowData(spe)$sigma.sq - rowData(spe)$trend
# ranking by difference above trend
rowData(spe)$rank_trend <- rank(-1 * rowData(spe)$difftrend, na.last = "keep")

# check for top genes
ix_svgs <- which(rowData(spe)$rank <= 10)
ix_svgs_trend <- which(rowData(spe)$rank_trend <= 10)
ix_hvgs <- which(rowData(spe)$rank_hvgs <= 10)

ix <- union(union(ix_svgs, ix_svgs_trend), ix_hvgs)
length(ix)
as.data.frame(rowData(spe)[ix, ])
```

```{r, fig.width=4.25, fig.height=4}
# plots
df <- as.data.frame(rowData(spe))


# properties of BRISC statistic

# BRISC statistic vs. mean logcounts
ix_nas <- is.na(df$sigma.sq)
ggplot(df[!ix_nas, ], aes(x = mean, y = sigma.sq)) + 
  geom_point() + 
  geom_line(aes(x = mean, y = trend), color = "blue", size = 1) + 
  labs(
    x = "mean logcounts", 
    y = "statistic (sigma.sq)", 
    title = "BRISC SVGs statistic"
  ) + 
  theme_bw()
ggsave("../plots/BRISC_stat_vs_logcounts.png", width = 5.5, height = 5)

# BRISC statistic rank vs. trend
which_zero <- max(df$rank_trend[!ix_nas][df$difftrend[!ix_nas] > 0])
ix_nas <- is.na(df$difftrend)
ggplot(df[!ix_nas, ], aes(x = rank_trend, y = difftrend)) + 
  geom_point() + 
  geom_vline(xintercept = 500, color = "red") + 
  geom_vline(xintercept = which_zero, color = "blue") + 
  labs(
    x = "rank (difference vs. trend)", 
    y = "statistic (difference vs. trend)", 
    title = "BRISC SVGs statistic"
  ) + 
  theme_bw()
ggsave("../plots/BRISC_difftrend_vs_rank.png", width = 5.5, height = 5)


# comparison with HVGs

# BRISC statistic vs. HVGs statistic
ix_nas <- is.na(df$sigma.sq) | is.na(df$hvgs_bio)
ggplot(df[!ix_nas, ], aes(x = hvgs_bio, y = difftrend)) + 
  geom_point() + 
  labs(
    x = "HVGs statistic (biological variation)", 
    y = "BRISC statistic (difference vs. trend)", 
    title = "SVGs and HVGs"
  ) + 
  theme_bw()
ggsave("../plots/BRISC_difftrend_vs_HVGs_bio.png", width = 5.5, height = 5)

# BRISC rank vs. HVGs rank
ix_nas <- is.na(df$rank_trend) | is.na(df$rank_hvgs)
ggplot(df[!ix_nas, ], aes(x = rank_hvgs, y = rank_trend)) + 
  geom_point() + 
  xlim(c(0, 1000)) + 
  ylim(c(0, 1000)) + 
  labs(
    x = "HVGs rank (biological variation) (top 1000)", 
    y = "BRISC rank (difference vs. trend) (top 1000)", 
    title = "SVGs and HVGs"
  ) + 
  theme_bw()
ggsave("../plots/BRISC_rank_vs_HVGs_rank.png", width = 5.5, height = 5)

# BRISC rank vs. HVGs rank
ix_nas <- is.na(df$rank_trend) | is.na(df$rank_hvgs)
ix_svgs_trend <- which(rowData(spe)$rank_trend <= 10)
ix_hvgs <- which(rowData(spe)$rank_hvgs <= 10)
ix_top <- union(ix_svgs_trend, ix_hvgs)
df_top <- df[ix_top, ]
ggplot(df[!ix_nas, ], aes(x = rank_hvgs, y = rank_trend)) + 
  geom_point() + 
  geom_point(data = df_top, aes(x = rank_hvgs, y = rank_trend), color = "maroon") + 
  geom_text_repel(data = df_top, mapping = aes(label = gene_name), color = "maroon") + 
  xlim(c(0, 30)) + 
  ylim(c(0, 30)) + 
  labs(
    x = "HVGs rank (biological variation) (top 30)", 
    y = "BRISC rank (difference vs. trend) (top 30)", 
    title = "SVGs and HVGs"
  ) + 
  theme_bw()
ggsave("../plots/BRISC_rank_vs_HVGs_rank_zoom.png", width = 5.5, height = 5)


# interesting genes from previous work

# BRISC statistic vs. HVGs statistic
ix_nas <- is.na(df$sigma.sq) | is.na(df$hvgs_bio)
ix_interesting <- which(df$gene_name %in% c("MOBP", "PCP4", "SNAP25", "HBB", "IGKC", "NPY"))
df_interesting <- df[ix_interesting, ]
ggplot(df[!ix_nas, ], aes(x = hvgs_bio, y = difftrend)) + 
  geom_point() + 
  geom_point(data = df_interesting, color = "blue") + 
  geom_text_repel(data = df_interesting, mapping = aes(label = gene_name), color = "blue") + 
  labs(
    x = "HVGs statistic (biological variation)", 
    y = "BRISC statistic (difference vs. trend)", 
    title = "SVGs and HVGs"
  ) + 
  theme_bw()
ggsave("../plots/BRISC_difftrend_vs_HVGs_bio_interesting.png", width = 5.5, height = 5)

# BRISC rank vs. HVGs rank
ix_nas <- is.na(df$rank_trend) | is.na(df$rank_hvgs)
ix_interesting <- which(df$gene_name %in% c("MOBP", "PCP4", "SNAP25", "HBB", "IGKC", "NPY"))
df_interesting <- df[ix_interesting, ]
ggplot(df[!ix_nas, ], aes(x = rank_hvgs, y = rank_trend)) + 
  geom_point() + 
  geom_point(data = df_interesting, color = "blue") + 
  geom_text_repel(data = df_interesting, mapping = aes(label = gene_name), color = "blue") + 
  xlim(c(0, 200)) + 
  ylim(c(0, 200)) + 
  labs(
    x = "HVGs rank (biological variation) (top 200)", 
    y = "BRISC rank (difference vs. trend) (top 200)", 
    title = "SVGs and HVGs"
  ) + 
  theme_bw()
ggsave("../plots/BRISC_rank_vs_HVGs_rank_interesting.png", width = 5.5, height = 5)
```


