---
title: "SVGs BRISC analyses: DLPFC subset with covariates"
author: "Lukas Weber"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Introduction

Analyses for SVGs BRISC method: human DLPFC dataset (subset) with covariates

Note: installation of MERINGUE and SPARK both require compilation

```{r, message=FALSE}
library(SpatialExperiment)
library(dplyr)
library(ggplot2)
```


## Load data

Load dataset and subset spots.

```{r, message=FALSE}
# load dataset
#library(STexampleData)
#spe <- Visium_humanDLPFC()

# temporary until new objects added to Bioc-devel
spe <- readRDS("../../../STexampleData/ExperimentHub_files/Bioc_3_14/Visium_humanDLPFC.rds")
dim(spe)
```

```{r}
# keep only spots in tissue
spe <- spe[, spatialData(spe)$in_tissue == 1]
dim(spe)

# subset small number of spots
n <- 200
set.seed(123)
ix <- sample(seq_len(ncol(spe)), 200)

spe <- spe[, ix]
dim(spe)
```

```{r}
# check
colData(spe)
spatialData(spe)
head(spatialCoords(spe))
```


## Preprocessing

Re-using code from OSTA

```{r, message=FALSE}
# QUALITY CONTROL (QC)

library(scater)
# identify mitochondrial genes
is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name)
table(is_mito)
# spot-level filtering
spe <- addPerCellQC(spe, subsets = list(mito = is_mito))
qc_lib_size <- colData(spe)$sum < 500
qc_detected <- colData(spe)$detected < 250
qc_mito <- colData(spe)$subsets_mito_percent > 30
qc_cell_count <- colData(spe)$cell_count > 12
# filter low-quality spots
colData(spe)$discard <- qc_lib_size | qc_detected | qc_mito | qc_cell_count
table(colData(spe)$discard)
spe <- spe[, !colData(spe)$discard]
dim(spe)

# NORMALIZATION

library(scran)
# calculate size factors
spe <- computeSumFactors(spe)
# calculate logcounts (log-transformed normalized counts)
spe <- logNormCounts(spe)

# FEATURE SELECTION

# fit mean-variance relationship
dec <- modelGeneVar(spe)
# select top HVGs
top_hvgs <- getTopHVGs(dec, prop = 0.1)
str(top_hvgs)
# keep HVG statistics
rowData(spe) <- DataFrame(left_join(
  as.data.frame(rowData(spe)), as.data.frame(cbind(dec, gene_id = rownames(dec))), 
  by = "gene_id"
))
head(rowData(spe))

# keep only HVGs for now
spe <- spe[top_hvgs, ]
dim(spe)
```


## Run SVGs methods

### Run BRISC

Using parallelized wrapper function in `spatzli` package.

notes:
- runtime for bootstrapping to get confidence intervals is ~100x slower
- to do: add slower version to rank SVGs by significance of `sigma.sq` parameter (using bootstrap confidence intervals)

```{r}
library(spatzli)
```

```{r}
# run BRISC to calculate statistics for each gene
runtime_brisc <- system.time({
  spe_base <- rankSVGsBRISC(spe, n_threads = 4)
})

runtime_brisc
```


### Run BRISC with covariates

Include covariates for manually annotated ground truth layers

```{r}
# run BRISC to calculate statistics for each gene
runtime_brisc_covariates <- system.time({
  spe_covariates <- rankSVGsBRISC(spe, x = model.matrix(~ colData(spe)$ground_truth), n_threads = 4)
})

runtime_brisc_covariates
```


### Compare gene lists

```{r}
# calculate ranks

# BRISC
rowData(spe_base)$rank_svgs_sigmasq <- rank(-1 * rowData(spe_base)$sigma.sq, na.last = "keep")
rowData(spe_covariates)$rank_svgs_sigmasq <- rank(-1 * rowData(spe_covariates)$sigma.sq, na.last = "keep")

# HVGs
rowData(spe_base)$rank_hvgs_bio <- rank(-1 * rowData(spe_base)$bio, na.last = "keep")
rowData(spe_covariates)$rank_hvgs_bio <- rank(-1 * rowData(spe_covariates)$bio, na.last = "keep")
```

```{r}
# top genes
rowData(spe_base)$gene_name[rowData(spe_base)$rank_svgs_sigmasq <= 10]
rowData(spe_covariates)$gene_name[rowData(spe_covariates)$rank_svgs_sigmasq <= 10]
```


### BRISC vs. non-spatial HVGs

```{r, fig.width=4.25, fig.height=4}
ggplot(as.data.frame(rowData(spe_base)), 
       aes(x = bio, y = sigma.sq)) + 
  geom_point(pch = 1) + 
  geom_smooth() + 
  labs(x = "HVGs: bio", 
       y = "BRISC: sigma.sq") + 
  ggtitle("Statistics") + 
  theme_bw()
```


### Plot top SVGs

```{r}
# identify top SVGs
ix_top_brisc_base <- which.max(rowData(spe_base)$sigma.sq)
ix_top_brisc_covariates <- which.max(rowData(spe_covariates)$sigma.sq)
```

```{r, fig.width=5, fig.height=4}
ggplot(as.data.frame(cbind(spatialCoords(spe_base), 
                           counts = counts(spe_base)[ix_top_brisc_base, ])), 
       aes(x = x, y = y, color = counts)) + 
  coord_fixed() + 
  geom_point(pch = 16, size = 2.5) + 
  scale_y_reverse() + 
  scale_color_gradient(low = "gray90", high = "blue") + 
  ggtitle(paste0("BRISC (base): ", rowData(spe)$gene_name[ix_top_brisc_base])) + 
  theme_bw()
```

```{r, fig.width=5, fig.height=4}
ggplot(as.data.frame(cbind(spatialCoords(spe_covariates), 
                           counts = counts(spe_covariates)[ix_top_brisc_covariates, ])), 
       aes(x = x, y = y, color = counts)) + 
  coord_fixed() + 
  geom_point(pch = 16, size = 2.5) + 
  scale_y_reverse() + 
  scale_color_gradient(low = "gray90", high = "blue") + 
  ggtitle(paste0("BRISC (covariates): ", rowData(spe)$gene_name[ix_top_brisc_covariates])) + 
  theme_bw()
```


### Plot layers

Plot ground truth layer labels for reference

```{r, fig.width=5, fig.height=4}
ggplot(as.data.frame(cbind(spatialCoords(spe_base), colData(spe))), 
       aes(x = x, y = y, color = ground_truth)) + 
  coord_fixed() + 
  scale_y_reverse() + 
  geom_point(pch = 16, size = 2.5) + 
  ggtitle(paste0("Ground truth layers")) + 
  theme_bw()
```


## Session info

```{r}
sessionInfo()
```

