---
title: "Analyses SVGs using BRISC: human DLPFC"
author: "Lukas Weber"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Introduction

Analyses for SVGs using BRISC methodology, spatzli wrapper, human DLPFC dataset.

```{r, message = FALSE}
library(SpatialExperiment)
library(STexampleData)
library(spatzli)
```


## Load data

```{r}
spe <- Visium_humanDLPFC()
dim(spe)
assayNames(spe)
```


## Preprocessing

Preprocessing steps: filter low-expressed genes, filter mitochondrial genes, normalization, and logcounts. Using convenience function in `spatzli` package.

```{r}
spe <- preprocessSVGs(spe)
dim(spe)
assayNames(spe)
```


## Identify SVGs

Run SVGs method using BRISC, parallelized implementation in `spatzli` package.

```{r}
runtime_brisc <- system.time({
  spe <- runSVGsBRISC(spe, x = NULL, n_threads = 4)
})
```

```{r}
runtime_brisc
```

```{r}
file <- "../outputs/spe_brisc_spatzli.rds"
saveRDS(spe, file = file)
```


## Session info

```{r}
sessionInfo()
```

