---
title: "Comparison of SVGs methods: Slide-seqV2 hippocampus"
author: "Lukas Weber"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Introduction

Analyses for SVGs: using Slide-seqV2 dataset from mouse hippocampus (from Stickels et al. 2021).

```{r, message = FALSE}
library(SpatialExperiment)
library(spatzli)
library(ggspavis)
library(readr)
library(ggplot2)
```


# Load data

Load Slide-seqV2 dataset from mouse hippocampus (from Stickels et al. 2021) and create `SpatialExperiment` object.

Notes:

- raw data is stored as a `.txt.gz` file containing a column of gene IDs and expression matrix
- too large to store in memory on laptop, so need to run this on JHPCE


```{r, eval = FALSE}
dir_data <- "../../data/Slide_seqV2_hippo/raw_data"

# load expression matrix
# runtime: several minutes
exprs <- read_table(file.path(dir_data, "Puck_200115_08.digital_expression.txt.gz"))
dim(exprs)
format(object.size(exprs), units = "GB")
exprs[1:6, 1:6]

# get gene IDs from first column
gene_ids <- exprs$GENE
str(gene_ids)
length(gene_ids)

# get barcode IDs from column names (excluding gene IDs column)
bead_ids <- colnames(exprs)[-1]
str(bead_ids)
length(bead_ids)

# convert expression matrix to numeric matrix without gene IDs
exprs <- exprs[, -1]
exprs <- as.matrix(exprs)
stopifnot(nrow(exprs) == length(gene_ids))
rownames(exprs) <- gene_ids

dim(exprs)
format(object.size(exprs), units = "GB")
exprs[1:6, 1:6]
```

```{r, eval = FALSE}
# load beads info
# note: contains mix of tab-delimited and comma-delimited
file <- file.path(dir_data, "Puck_200115_08_bead_locations.csv")
bead_locations_colnames <- unlist(strsplit(readLines(file, n = 1), "\t"))
bead_locations <- read_csv(file, skip = 1, col_names = FALSE)
colnames(bead_locations) <- bead_locations_colnames
dim(bead_locations)
head(bead_locations)

stopifnot(nrow(bead_locations) == ncol(exprs))
stopifnot(all(bead_ids == bead_locations$barcodes))
```

```{r, eval = FALSE}
# create SpatialExperiment
row_data <- DataFrame(
  gene_name = gene_ids
)

col_data <- DataFrame(
  barcode_id = bead_ids
)

spatial_data <- DataFrame(
  barcode_id = bead_ids
)

spatial_coords <- as.matrix(bead_locations[, c("xcoord", "ycoord")])
rownames(spatial_coords) <- bead_ids

exprs_sparse <- as(exprs, "dgCMatrix")

spe <- SpatialExperiment(
  rowData = row_data, 
  colData = col_data, 
  spatialCoords = spatial_coords, 
  spatialData = spatial_data, 
  assays = list(counts = exprs_sparse)
)

spe

format(object.size(spe), units = "MB")
```

```{r, eval = FALSE}
# save SpatialExperiment object
dir_out <- "../../data/Slide_seqV2_hippo/SPE"
file_out <- file.path(dir_out, "Slide_seqV2_hippo_SPE.rds")
saveRDS(spe, file = file_out)
```

```{r}
# alternatively: load saved object
dir_out <- "../../data/Slide_seqV2_hippo/SPE"
file_out <- file.path(dir_out, "Slide_seqV2_hippo_SPE.rds")
spe <- readRDS(file_out)
spe
```


---

# Plot known genes

Plot expression of some known cell type markers and known spatially variable genes (SVGs) from Cable et al. (2021), Figure 6.

```{r, fig.width = 7, fig.height = 6}
# known cell type markers
marker_names <- c("Snap25")
ix_markers <- match(marker_names, rowData(spe)$gene_name)
ix_markers

# known SVGs
svg_names <- c("Rgs14", "Cpne9")
ix_svgs <- match(svg_names, rowData(spe)$gene_name)
ix_svgs

names_all <- c(marker_names, svg_names)
ix_all <- c(ix_markers, ix_svgs)

# plot expression
df <- as.data.frame(cbind(colData(spe), spatialCoords(spe)))

for (i in seq_along(ix_all)) {
  df$sum <- counts(spe)[ix_all[i], ]
  
  p <- ggplot(df, aes_string(x = "xcoord", y = "ycoord", color = "sum")) + 
    geom_point(size = 0.1) + 
    scale_color_gradient2(low = "gray97", mid = "red", high = "darkred", 
                          midpoint = mean(range(df$sum))) + 
    coord_fixed() + 
    ggtitle(names_all[i]) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


---

# Session info

```{r}
sessionInfo()
```

