---
title: "Analyses SVGs: human DLPFC"
author: "Lukas Weber"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Introduction

Analyses for SVGs: comparison of methods including BRISC, GpGp, HVGs, using spatzli wrappers, human DLPFC dataset.

```{r, message = FALSE}
library(SpatialExperiment)
library(STexampleData)
library(spatzli)
library(scran)
library(dplyr)
library(ggplot2)
```


## Load data

```{r}
spe <- Visium_humanDLPFC()
dim(spe)
assayNames(spe)
```


## Preprocessing

Preprocessing steps: filter low-expressed genes, filter mitochondrial genes, normalization, and logcounts. Using convenience function in `spatzli` package.

```{r}
spe <- preprocessSVGs(spe)
dim(spe)
assayNames(spe)
```


## Identify SVGs

### HVGs (non-spatial)

Non-spatial HVGs for baseline comparison.

```{r}
# fit mean-variance relationship
dec <- modelGeneVar(spe)

# store in object
stopifnot(all(rownames(dec) == rowData(spe)$gene_id))
dec_keep <- dec[, c("bio", "p.value", "FDR")]
colnames(dec_keep) <- c("bio", "hvgs_pval", "hvgs_fdr")
rowData(spe) <- cbind(rowData(spe), dec_keep)
```


### BRISC

To do: return log-likelihoods from BRISC fork to enable likelihood ratio tests.


### GpGp

Run SVGs method using GpGp, and calculate likelihood ratio tests comparing models with and without spatial terms.

Runtime: ~5 hours using 4 cores on laptop.

Note: either run code in first couple of code blocks below (5 hours runtime) or load pre-saved results object in next code block.

```{r}
# runtime_gpgp <- system.time({
#   spe_gpgp <- runSVGsGpGp(spe, x = NULL, n_threads = 4)
# })
```

```{r}
# runtime_gpgp
```

```{r}
# file <- "../outputs/spe_gpgp_spatzli.rds"
# saveRDS(spe_gpgp, file = file)
```

```{r}
# alternatively: load previously saved results
spe_gpgp <- readRDS("../outputs/spe_gpgp_spatzli.rds")
```


## Comparison

### Plots

**Set up data frame for plotting**

```{r}
# data frame for plotting
stopifnot(all(rowData(spe)$gene_id == rowData(spe_gpgp)$gene_id))
dim(rowData(spe))
dim(rowData(spe_gpgp))

df <- full_join(as.data.frame(rowData(spe)), as.data.frame(rowData(spe_gpgp)))
dim(df)
```

---

**GpGp statistic vs. HVGs statistic**

Using log(1 + x) transformation for improved visibility.

```{r, fig.width = 5, fig.height = 5}
ggplot(df, aes(x = bio, y = sigmasq)) + 
  geom_point(pch = 1) + 
  geom_smooth() + 
  scale_x_continuous(trans = "log1p") + 
  scale_y_continuous(trans = "log1p") + 
  labs(x = "HVGs statistic (bio)", 
       y = "GpGp statistic (sigmasq)") + 
  ggtitle("GpGp statistic vs. HVGs statistic") + 
  theme_bw()
```

---

**Significance (adjusted p-values) vs. effect size (proportion spatial variance) for GpGp method**

```{r, fig.width = 5, fig.height = 5}
ggplot(df, aes(x = prop_sv, y = padj)) + 
  geom_point(pch = 1) + 
  labs(x = "proportion spatial variance", 
       y = "adjusted p-value") + 
  ggtitle("Significance vs. effect size: GpGp method") + 
  theme_bw()
```

---

**Effect sizes (proportion spatial variance) for genes with p-values exactly equal to zero for GpGp method**

```{r, fig.width = 5, fig.height = 4}
# number of genes with p-value == 0
table(df$padj == 0)

df_sub <- df[df$padj == 0, ]
dim(df_sub)

ggplot(df_sub, aes(x = prop_sv)) + 
  geom_histogram(bins = 40, color = "black", fill = "navy") + 
  geom_vline(xintercept = 0.2, color = "red") + 
  labs(x = "proportion spatial variance") + 
  ggtitle("Effect sizes for genes with p-values == 0: GpGp") + 
  theme_bw()
```


### Significance rule

**Example of overall significance rule using both statistical significance and effect size. Six known interesting genes in this dataset (SNAP25, PCP4, MOBP, HBB, IGKC, NPY) are all detected by this rule.**

```{r, fig.width = 5, fig.height = 4}
# number of genes with p-value == 0
table(df$padj == 0)
# number of genes with effect size >= 0.2
table(df$prop_sv >= 0.2)

# number of genes with 'p-value == 0' & 'effect size >= 0.2'
table(df$padj == 0 & df$prop_sv >= 0.2)
# contingency table
table(df$padj == 0, df$prop_sv >= 0.2)

# print gene names
sig <- df$padj == 0 & df$prop_sv >= 0.2
sig_names <- rowData(spe)$gene_name[sig]
sig_names

# check known interesting genes in this dataset
interesting <- c("SNAP25", "PCP4", "MOBP", "HBB", "IGKC", "NPY")
interesting %in% sig_names
```


## Session info

```{r}
sessionInfo()
```

