---
title: "SVGs BRISC analyses: subsetting HVGs"
author: "Lukas Weber"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 2
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Introduction

Analyses for SVGs BRISC method - subsetting to HVGs.


## Preprocessing

Re-using code from OSTA: remove mitochondrial genes, select HVGs

```{r, message=FALSE}
# LOAD DATA

library(SpatialExperiment)
library(STexampleData)
spe <- Visium_humanDLPFC()

# QUALITY CONTROL (QC)

library(scater)
# subset to keep only spots over tissue
spe <- spe[, spatialData(spe)$in_tissue == 1]
# identify mitochondrial genes
is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name)
# calculate per-spot QC metrics
spe <- addPerCellQC(spe, subsets = list(mito = is_mito))
# select QC thresholds
qc_lib_size <- colData(spe)$sum < 500
qc_detected <- colData(spe)$detected < 250
qc_mito <- colData(spe)$subsets_mito_percent > 30
qc_cell_count <- colData(spe)$cell_count > 12
# combined set of discarded spots
discard <- qc_lib_size | qc_detected | qc_mito | qc_cell_count
colData(spe)$discard <- discard
# filter low-quality spots
spe <- spe[, !colData(spe)$discard]

# NORMALIZATION

library(scran)
# quick clustering for pool-based size factors
set.seed(123)
qclus <- quickCluster(spe)
# calculate size factors
spe <- computeSumFactors(spe, cluster = qclus)
# calculate logcounts (log-transformed normalized counts)
spe <- logNormCounts(spe)

# FEATURE SELECTION

# remove mitochondrial genes
spe <- spe[!is_mito, ]
# fit mean-variance relationship
dec <- modelGeneVar(spe)
# select top HVGs
top_hvgs <- getTopHVGs(dec, prop = 0.1)
```

```{r, message=FALSE}
# subset object to contain HVGs only
spe <- spe[rowData(spe)$gene_id %in% top_hvgs, ]
dim(spe)

# include HVG ranks
rowData(spe)$rank_hvgs <- match(rowData(spe)$gene_id, top_hvgs)

# include HVG statistic (in correct order)
bio <- dec$bio
names(bio) <- rownames(dec)
head(bio)
rowData(spe)$bio <- bio[rownames(rowData(spe))]
```


## Run BRISC to rank SVGs

Rank SVGs (within set of HVGs) using BRISC - using implementation in `spatzli` package

runtime: approx. 11 minutes (with 4 cores on laptop; after subsetting to HVGs)

(note: runtime for bootstrapping to get confidence intervals is approx. 100 times slower - so would expect ~10 hours)

to do: include slower version to rank SVGs by significance of `sigma.sq` parameter (using bootstrap confidence intervals)

```{r}
# run BRISC to calculate statistics for each gene
library(spatzli)

runtime <- system.time({
  spe <- rankSVGsBRISC(spe, n_threads = 4)
})
```

```{r}
# rank SVGs by sigma.sq parameter
rowData(spe)$rank_svgs_sigmasq <- rank(-1 * rowData(spe)$sigma.sq)

# rank SVGs by fraction of spatial variance (FSV)
rowData(spe)$rank_svgs_fsv <- rank(-1 * rowData(spe)$fsv)
```


## Comparisons

Comparisons: SVG using `sigma.sq` parameter, SVGs using `FSV`

```{r, message=FALSE}
library(ggplot2)
```


```{r}
# calculate total UMIs per gene
rowData(spe)$sum <- rowSums(counts(spe))
# calculate mean logcounts per gene
rowData(spe)$mean <- rowMeans(logcounts(spe))
```


## Compare rankings

```{r, fig.width=4.25, fig.height=4}
ggplot(as.data.frame(rowData(spe)), aes(x = rank_hvgs, y = rank_svgs_sigmasq)) + 
  geom_point(pch = 1) + 
  theme_bw()
ggsave("../plots_new/rankSVGssigmasq_rankHVGs.png", width = 5.5, height = 5)

ggplot(as.data.frame(rowData(spe)), aes(x = rank_hvgs, y = rank_svgs_fsv)) + 
  geom_point(pch = 1) + 
  theme_bw()
ggsave("../plots_new/rankSVGsFSV_rankHVGs.png", width = 5.5, height = 5)

ggplot(as.data.frame(rowData(spe)), aes(x = rank_svgs_sigmasq, y = rank_svgs_fsv)) + 
  geom_point(pch = 1) + 
  theme_bw()
ggsave("../plots_new/rankSVGsFSV_rankSVGssigmasq.png", width = 5.5, height = 5)
```


## Compare statistics

```{r, fig.width=4.25, fig.height=4}
ggplot(as.data.frame(rowData(spe)), aes(x = bio, y = sigma.sq)) + 
  geom_point(pch = 1) + 
  theme_bw()
ggsave("../plots_new/sigmasq_bio.png", width = 5.5, height = 5)

ggplot(as.data.frame(rowData(spe)), aes(x = bio, y = fsv)) + 
  geom_point(pch = 1) + 
  theme_bw()
ggsave("../plots_new/fsv_bio.png", width = 5.5, height = 5)
```

